---
title: 'Reproducible Research: Peer Assessment 2'
subtitle: by Henrique Souza (github.com/htssouza)
output:
  html_document:
    keep_md: yes
    pandoc_args: [
      "+RTS", "-K64m",
      "-RTS"
    ]
        
---

# Data Processing

Downloading and loading the data:

```{r echo = TRUE} 
storm_data_url <- 
    "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(storm_data_url, destfile = "stormdata.csv.bz2", method = "curl")
storm_data <- read.csv(bzfile("stormdata.csv.bz2"))
```

Sample data:

```{r echo = TRUE} 
str(storm_data)
```

Considering only (valid) states of United States:

```{r echo = TRUE} 
us_states_and_dc <- 
c("DC", "AL","MT","AK","NE","AZ","NV","AR","NH","CA","NJ","CO","NM","CT","NY",
"DE","NC","FL","ND","GA","OH","HI","OK","ID","OR","IL","PA","IN","RI",
"IA","SC","KS","SD","KY","TN","LA","TX","ME","UT","MD","VT","MA","VA",
"MI","WA","MN","WV","MS","WI","MO","WY")

storm_data <- storm_data[storm_data$STATE %in% us_states_and_dc,]
```

Converting dates columns:

```{r echo = TRUE} 
storm_data$BGN_DATE <- as.Date(storm_data$BGN_DATE, "%m/%d/%Y")
storm_data$END_DATE <- as.Date(storm_data$END_DATE, "%m/%d/%Y")
```

Normalizing Property damage:

```{r echo = TRUE} 

storm_data$PROPDMG[storm_data$PROPDMGEXP == "K"] <-
    (storm_data$PROPDMG[storm_data$PROPDMGEXP == "K"] * 1000)
storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == "K"] <- ""

storm_data$PROPDMG[storm_data$PROPDMGEXP == "k"] <-
    (storm_data$PROPDMG[storm_data$PROPDMGEXP == "k"] * 1000)
storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == "k"] <- ""

storm_data$PROPDMG[storm_data$PROPDMGEXP == "M"] <-
    (storm_data$PROPDMG[storm_data$PROPDMGEXP == "M"] * 1000 * 1000)
storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == "M"] <- ""

storm_data$PROPDMG[storm_data$PROPDMGEXP == "m"] <-
    (storm_data$PROPDMG[storm_data$PROPDMGEXP == "m"] * 1000 * 1000)
storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == "m"] <- ""

storm_data$PROPDMG[storm_data$PROPDMGEXP == "B"] <-
    (storm_data$PROPDMG[storm_data$PROPDMGEXP == "B"] * 1000 * 1000 * 1000)
storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == "B"] <- ""

storm_data$PROPDMG[storm_data$PROPDMGEXP == "b"] <-
    (storm_data$PROPDMG[storm_data$PROPDMGEXP == "b"] * 1000 * 1000 * 1000)
storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == "b"] <- ""

for(exp in as.character(1:10)) {
    storm_data$PROPDMG[storm_data$PROPDMGEXP == exp] <-
        (storm_data$PROPDMG[storm_data$PROPDMGEXP == exp] * (10 ^ as.integer(exp)))
    storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == exp] <- ""
}

```

Normalizing Crop damage:

```{r echo = TRUE} 

storm_data$CROPDMG[storm_data$CROPDMGEXP == "K"] <-
    (storm_data$CROPDMG[storm_data$CROPDMGEXP == "K"] * 1000)
storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == "K"] <- ""

storm_data$CROPDMG[storm_data$CROPDMGEXP == "k"] <-
    (storm_data$CROPDMG[storm_data$CROPDMGEXP == "k"] * 1000)
storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == "k"] <- ""

storm_data$CROPDMG[storm_data$CROPDMGEXP == "M"] <-
    (storm_data$CROPDMG[storm_data$CROPDMGEXP == "M"] * 1000 * 1000)
storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == "M"] <- ""

storm_data$CROPDMG[storm_data$CROPDMGEXP == "m"] <-
    (storm_data$CROPDMG[storm_data$CROPDMGEXP == "m"] * 1000 * 1000)
storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == "m"] <- ""

storm_data$CROPDMG[storm_data$CROPDMGEXP == "B"] <-
    (storm_data$CROPDMG[storm_data$CROPDMGEXP == "B"] * 1000 * 1000 * 1000)
storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == "B"] <- ""

storm_data$CROPDMG[storm_data$CROPDMGEXP == "b"] <-
    (storm_data$CROPDMG[storm_data$CROPDMGEXP == "b"] * 1000 * 1000 * 1000)
storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == "b"] <- ""

for(exp in as.character(1:10)) {
    storm_data$CROPDMG[storm_data$CROPDMGEXP == exp] <-
        (storm_data$CROPDMG[storm_data$CROPDMGEXP == exp] * (10 ^ as.integer(exp)))
    storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == exp] <- ""
}

```


Considering the data with valid property and crop manage:

```{r echo = TRUE} 
storm_data <- storm_data[storm_data$PROPDMGEXP == "",]
storm_data <- storm_data[storm_data$CROPDMGEXP == "",]
```

Computing population harm per state and event type:

```{r echo = TRUE} 
pop_harm_per_evtype <- aggregate(
    storm_data[c("FATALITIES", "INJURIES")],
    by=storm_data[c("STATE", "EVTYPE")],
    FUN=sum)

pop_harm_per_evtype <-
    pop_harm_per_evtype [
        (pop_harm_per_evtype$FATALITIES +
        pop_harm_per_evtype$INJURIES)  > 0
    ,]

pop_harm_per_evtype <- pop_harm_per_evtype[
    order(
        pop_harm_per_evtype$STATE,
        -pop_harm_per_evtype$FATALITIES,
        -pop_harm_per_evtype$INJURIES
    ),
]

```

Filtering the most population harmful event per state:

```{r echo = TRUE} 

pop_harm_per_evtype$temp <- (
    pop_harm_per_evtype$FATALITIES * 1000000 + 
    pop_harm_per_evtype$INJURIES
)

pop_harm_top_type_per_state <- aggregate(
    temp ~ STATE, pop_harm_per_evtype, max
)

pop_harm_top_type_per_state <- merge (
    pop_harm_top_type_per_state, pop_harm_per_evtype)

pop_harm_top_type_per_state <- pop_harm_top_type_per_state[,
    names(pop_harm_top_type_per_state) != "temp"
]

```

Computing economical (property and crop) damage per state and event type:

```{r echo = TRUE} 

storm_data$ECODMG <- (storm_data$PROPDMG + storm_data$CROPDMG)

eco_harm_per_evtype <- aggregate(
    storm_data["ECODMG"],
    by=storm_data[c("STATE", "EVTYPE")],
    FUN=sum)

eco_harm_per_evtype <-
    eco_harm_per_evtype [        
        eco_harm_per_evtype$ECODMG  > 0
    ,]

eco_harm_per_evtype <- eco_harm_per_evtype[
    order(
        eco_harm_per_evtype$STATE,
        -eco_harm_per_evtype$ECODMG
    ),
]

```

Filtering the most economical harmful event per state:

```{r echo = TRUE} 

eco_harm_top_type_per_state <- aggregate(
    ECODMG ~ STATE, eco_harm_per_evtype, max
)

eco_harm_top_type_per_state <- merge (
    eco_harm_top_type_per_state, eco_harm_per_evtype)

]

```




