---
title: 'The most harmful weather events accross United States '
subtitle: by Henrique Souza (github.com/htssouza)
output:
  html_document:
    keep_md: yes
    pandoc_args:
    - +RTS
    - -K512m
    - -RTS
---

# Synopsis

In this report we will indicate the most harmful weather
events accross Unites States.

This information will be useful in order to concentrate
the efforts to mitigate and prevent damages caused by 
these severe events.

# Data Processing

Loading dependency libraries:

```{r echo = TRUE} 
library(ggplot2)
```

Sets locale:

```{r echo = TRUE} 
Sys.setlocale("LC_ALL", "en_US")
```

Downloading and loading the data:

```{r echo = TRUE, cache = TRUE} 
storm_data_url <- 
    "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(storm_data_url, destfile = "stormdata.csv.bz2", method = "curl")
storm_data <- read.csv(bzfile("stormdata.csv.bz2"))
```

Sample data:

```{r echo = TRUE} 
str(storm_data)
```

Considering only (valid) states of United States:

```{r echo = TRUE} 
us_states_and_dc <- 
c("DC", "AL","MT","AK","NE","AZ","NV","AR","NH","CA","NJ","CO","NM","CT","NY",
"DE","NC","FL","ND","GA","OH","HI","OK","ID","OR","IL","PA","IN","RI",
"IA","SC","KS","SD","KY","TN","LA","TX","ME","UT","MD","VT","MA","VA",
"MI","WA","MN","WV","MS","WI","MO","WY")

us_states_and_dc

storm_data <- storm_data[storm_data$STATE %in% us_states_and_dc,]
```

Converting dates columns:

```{r echo = TRUE} 
storm_data$BGN_DATE <- as.Date(storm_data$BGN_DATE, "%m/%d/%Y")
storm_data$END_DATE <- as.Date(storm_data$END_DATE, "%m/%d/%Y")
storm_data$YEAR <- as.integer(format(storm_data$BGN_DATE, "%Y"))
```

Normalizing Property damage:

```{r echo = TRUE} 

storm_data$PROPDMG[storm_data$PROPDMGEXP == "K"] <-
    (storm_data$PROPDMG[storm_data$PROPDMGEXP == "K"] * 1000)
storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == "K"] <- ""

storm_data$PROPDMG[storm_data$PROPDMGEXP == "k"] <-
    (storm_data$PROPDMG[storm_data$PROPDMGEXP == "k"] * 1000)
storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == "k"] <- ""

storm_data$PROPDMG[storm_data$PROPDMGEXP == "M"] <-
    (storm_data$PROPDMG[storm_data$PROPDMGEXP == "M"] * 1000 * 1000)
storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == "M"] <- ""

storm_data$PROPDMG[storm_data$PROPDMGEXP == "m"] <-
    (storm_data$PROPDMG[storm_data$PROPDMGEXP == "m"] * 1000 * 1000)
storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == "m"] <- ""

storm_data$PROPDMG[storm_data$PROPDMGEXP == "B"] <-
    (storm_data$PROPDMG[storm_data$PROPDMGEXP == "B"] * 1000 * 1000 * 1000)
storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == "B"] <- ""

storm_data$PROPDMG[storm_data$PROPDMGEXP == "b"] <-
    (storm_data$PROPDMG[storm_data$PROPDMGEXP == "b"] * 1000 * 1000 * 1000)
storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == "b"] <- ""

for(exp in as.character(1:10)) {
    storm_data$PROPDMG[storm_data$PROPDMGEXP == exp] <-
        (storm_data$PROPDMG[storm_data$PROPDMGEXP == exp] * (10 ^ as.integer(exp)))
    storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == exp] <- ""
}

```

Normalizing Crop damage:

```{r echo = TRUE} 

storm_data$CROPDMG[storm_data$CROPDMGEXP == "K"] <-
    (storm_data$CROPDMG[storm_data$CROPDMGEXP == "K"] * 1000)
storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == "K"] <- ""

storm_data$CROPDMG[storm_data$CROPDMGEXP == "k"] <-
    (storm_data$CROPDMG[storm_data$CROPDMGEXP == "k"] * 1000)
storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == "k"] <- ""

storm_data$CROPDMG[storm_data$CROPDMGEXP == "M"] <-
    (storm_data$CROPDMG[storm_data$CROPDMGEXP == "M"] * 1000 * 1000)
storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == "M"] <- ""

storm_data$CROPDMG[storm_data$CROPDMGEXP == "m"] <-
    (storm_data$CROPDMG[storm_data$CROPDMGEXP == "m"] * 1000 * 1000)
storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == "m"] <- ""

storm_data$CROPDMG[storm_data$CROPDMGEXP == "B"] <-
    (storm_data$CROPDMG[storm_data$CROPDMGEXP == "B"] * 1000 * 1000 * 1000)
storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == "B"] <- ""

storm_data$CROPDMG[storm_data$CROPDMGEXP == "b"] <-
    (storm_data$CROPDMG[storm_data$CROPDMGEXP == "b"] * 1000 * 1000 * 1000)
storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == "b"] <- ""

for(exp in as.character(1:10)) {
    storm_data$CROPDMG[storm_data$CROPDMGEXP == exp] <-
        (storm_data$CROPDMG[storm_data$CROPDMGEXP == exp] * (10 ^ as.integer(exp)))
    storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == exp] <- ""
}

```

Considering the data with valid property and crop manage:

```{r echo = TRUE} 
storm_data <- storm_data[storm_data$PROPDMGEXP == "",]
storm_data <- storm_data[storm_data$CROPDMGEXP == "",]
```

Computing population harm per state and event type:

```{r echo = TRUE} 
pop_harm_per_evtype <- aggregate(
    storm_data[c("FATALITIES", "INJURIES")],
    by=storm_data[c("STATE", "EVTYPE")],
    FUN=sum)

pop_harm_per_evtype <-
    pop_harm_per_evtype [
        (pop_harm_per_evtype$FATALITIES +
        pop_harm_per_evtype$INJURIES)  > 0
    ,]

pop_harm_per_evtype <- pop_harm_per_evtype[
    order(
        pop_harm_per_evtype$STATE,
        -pop_harm_per_evtype$FATALITIES,
        -pop_harm_per_evtype$INJURIES
    ),
]

```

Filtering the most population harmful event per state:

```{r echo = TRUE} 

pop_harm_per_evtype$temp <- (
    pop_harm_per_evtype$FATALITIES * 1000000 + 
    pop_harm_per_evtype$INJURIES
)

pop_harm_top_type_per_state <- aggregate(
    temp ~ STATE, pop_harm_per_evtype, max
)

pop_harm_top_type_per_state <- merge (
    pop_harm_top_type_per_state, pop_harm_per_evtype)

pop_harm_top_type_per_state <- pop_harm_top_type_per_state[,
    names(pop_harm_top_type_per_state) != "temp"
]

```

Computing populational (fatalities) damage per year:

```{r echo = TRUE} 

pop_harm_per_year <- aggregate(
    storm_data["FATALITIES"],
    by=storm_data["YEAR"],
    FUN=sum)

```

Computing economical (property and crop) damage per state and event type:

```{r echo = TRUE} 

storm_data$ECODMG <- (storm_data$PROPDMG + storm_data$CROPDMG)

eco_harm_per_evtype <- aggregate(
    storm_data["ECODMG"],
    by=storm_data[c("STATE", "EVTYPE")],
    FUN=sum)

eco_harm_per_evtype <-
    eco_harm_per_evtype [        
        eco_harm_per_evtype$ECODMG  > 0
    ,]

eco_harm_per_evtype <- eco_harm_per_evtype[
    order(
        eco_harm_per_evtype$STATE,
        -eco_harm_per_evtype$ECODMG
    ),
]

```

Filtering the most economical harmful event per state:

```{r echo = TRUE} 

eco_harm_top_type_per_state <- aggregate(
    ECODMG ~ STATE, eco_harm_per_evtype, max
)

eco_harm_top_type_per_state <- merge (
    eco_harm_top_type_per_state, eco_harm_per_evtype)

```

Computing economical (property and crop) damage per year:

```{r echo = TRUE} 

storm_data$ECODMGINM <- (storm_data$ECODMG / (1000000))

eco_harm_per_year <- aggregate(
    storm_data["ECODMGINM"],
    by=storm_data["YEAR"],
    FUN=sum)

```

# Results

POPULATIONAL damage per year:

```{r echo = TRUE} 
qplot(
    YEAR,
    FATALITIES,
    data = pop_harm_per_year,
    xlab = "Year",
    ylab = "Fatalities",
    geom=c("line"))
```

The most harmful events per state in terms of POPULATIONAL impacts are:

```{r echo = TRUE} 
pop_harm_top_type_per_state
```

ECONOMICAL damage per year:

```{r echo = TRUE} 
qplot(
    YEAR,
    ECODMGINM,
    data = eco_harm_per_year,
    xlab = "Year",
    ylab = "Economical damage (USD millions)",
    geom=c("line"))
```

The most harmful events per state in terms of ECONOMICAL impacts are:

```{r echo = TRUE} 
eco_harm_top_type_per_state
```


